---
title: "fortranClustering HowTo"
author: "Kai Lochbihler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction and installation
The package fortranClustering delivers a basic example of how to use a fortran subroutine in R. It also comes with an equivalent implementation of the fortran subroutine in R code and a simple example data set.

To build and install the package on a linux maching make sure that the devtools package is installed on your system then load it with
```{r, LoadDT}
library(devtools)
```

Now you can install the fortranClustering package directly from github
#```{r, InstallFC}
#install_github("MartinRoth/fortranClustering",ref="FirstAttempt")
#```
and load it
```{r, LoadFC}
library(fortranClustering)
```

## How to include fortran code in R

```fortran
subroutine clustering(data2d,startID,thres,finID,numIDs,tcl,missval,nx,ny)
    
  implicit none
  ! input
  real(kind=8), intent(in) :: data2d(nx,ny) ! the input 2D field
  integer, intent(in)      :: nx,ny         ! input dimensions
  integer, intent(in)      :: startID       ! the first ID to use
  real(kind=8), intent(in) :: thres         ! minimum value for clustering  
  real(kind=8), intent(in) :: missval       ! value for missing data
  ! output
  integer, intent(out)     :: finID         ! the last used ID
  integer, intent(out)     :: numIDs        ! the total number of clusters/IDs
  integer, intent(out)     :: tcl(nx,ny)    ! the clustered field holding the IDs
  ! counters etc.
  integer                  :: x,y           ! for spatial dimenstions
  integer                  :: conx,cony     ! for spatial dims in nested loop
  integer                  :: i,tp          ! for misc.
  integer                  :: clID          ! for iterating clusters
  integer, allocatable     :: allIDs(:)     ! helper vector for re-assigning of IDs
  integer                  :: neighb(2)     ! temporary memory for neighbouring grid points
  logical                  :: mask(nx,ny)   ! logical field; true if value > thres

  ! initialize variables and arrays
  tcl=-999
  mask=.false.
  clID=startID
  numIDs=0

  ! mask values higher than threshold and if not missing value
  do y=1,ny
    do x=1,nx
      if(data2d(x,y)>thres .AND. data2d(x,y).ne.missval)mask(x,y)=.true.
    end do
  end do
  
  ! check if there are any gridpoints to cluster
  if(ANY(mask))then
  
    ! assign IDs to continous cells
    do y=1,ny
      do x=1,nx
        neighb=-999
        if(mask(x,y))then
          ! gather neighbouring IDs; left,up
          if(x.ne.1)  neighb(1)=tcl((x-1),y)
          if(y.ne.1)  neighb(2)=tcl(x,(y-1))
          ! check if there is NO cluster around the current pixel; create new one
          if(ALL(neighb==-999))then
            tcl(x,y)=clID
            clID=clID+1
            numIDs=numIDs+1
          else
            ! both neighbours are in the same cluster
            if(neighb(1)==neighb(2).and.neighb(1).ne.-999)then
              tcl(x,y)=neighb(1)
            end if
            ! both neighbors are in different clusters but none of them is (-999)
            if(neighb(1).ne.neighb(2) .and. neighb(1).ne.-999 .and. neighb(2).ne.-999)then
              numIDs=numIDs-1
              tcl(x,y)=MINVAL(neighb)
              ! update the existing higher cluster with the lowest neighbour
              do cony=1,ny
                do conx=1,nx
                  if(tcl(conx,cony)==MAXVAL(neighb))tcl(conx,cony)=MINVAL(neighb)
                end do
              end do
            end if
            ! both neighbors are in different clusters but ONE of them is empty(-999)
            if(neighb(1).ne.neighb(2) .and. (neighb(1)==-999 .or. neighb(2)==-999))then
              tcl(x,y)=MAXVAL(neighb)
            end if
          end if
        end if
      end do
    end do
  
    ! gather IDs and rename to gapless ascending IDs
    if(numIDs>0)then
      allocate(allIDs(numIDs))
      allIDs=-999
      clID=startID-1
      tp=1
      do y=1,ny
        do x=1,nx
          if(.NOT.ANY(allIDs==tcl(x,y)) .AND. tcl(x,y).ne.-999)then
            allIDs(tp)=tcl(x,y)
            tp=tp+1
          end if
        end do
      end do
  
      do i=1,tp-1
        clID=clID+1
        do y=1,ny
          do x=1,nx
            if(tcl(x,y)==allIDs(i))then
              tcl(x,y)=clID
            end if
          end do
        end do
      end do
      deallocate(allIDs)
    end if
    
  end if
  ! return final cluster ID
  finID=clID
end subroutine clustering
```

```bash
R CMD SHLIB clustering2d.f90
```


```r
clusteringf90 <- function(x,startID=1,thres=0) {
  # only allow a matrix as input
  if(!is.matrix(x)){
    stop("x is not a matrix!")
  }
  # load the fortran subroutine
  if(!is.loaded('clustering2d')){
    dyn.load('clustering2d.so')
  }
  # set NA to -999
  x[is.na(x)] <- -999.0
  # get dimensions of input matrix
  nx <- dim(x)[1]
  ny <- dim(x)[2]
  # call the fortran subroutine
  retdata <- .Fortran("clustering",
                       data2d = as.double(x),
                       startID = as.integer(startID),
                       thres = as.double(thres),
                       finID = as.integer(1),
                       numIDs = as.integer(1),
                       tcl = matrix(rep(as.integer(1),nx*ny),ncol=nx),
                       missval=as.double(-999.0),
                       nx=as.integer(nx),
                       ny=as.integer(ny))
  # set -999 in tcl back to NA
  retdata$tcl[retdata$tcl==-999] <- NA
  # return the data
  return(retdata)
}
```


```{r, LoadData}
load("../sample_precip_data2d.gz")
sample_data <- precip.matrix
rm(precip.matrix)
sample_data[sample_data<0.000166666666667] <- NA
filled.contour(sample_data*60*10,color.palette=rainbow)

```


```{r, rspeed}
system.time({
  sample_output.f90 <- clusteringf90(sample_data)
})

# now the R implementation
system.time({
  sample_output.R <- clusteringR(sample_data)
})
```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
